\documentclass[a4paper,12pt]{article}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{mathtools}

\title{Symmetry Protection of Edge States in a 2D Topological Insulator}
\author{Amin Marie Ann, Azarcon Vince,Balangue Jabbar}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
  Investigate how time-reversal symmetry protects the gapless edge states in a 2D $Z_2$ topological insulator (like the Bernevig-Hughes-Zhang (BHZ) model).
\end{abstract}

\section{Mathematical Derivation}
\begin{enumerate}
    \item Start with the effective four-band BHZ model Hamiltonian in 2D:
    
    \[
    H(\mathbf{k}) = \begin{pmatrix}
    m(\mathbf{k}) & A k_x & 0 & -A k_y \\
    A k_x & -m(\mathbf{k}) & -A k_y & 0 \\
    0 & -A k_y & -m(\mathbf{k}) & -A k_x \\
    -A k_y & 0 & -A k_x & m(\mathbf{k})
    \end{pmatrix},
    \]
    where $m(\mathbf{k}) = M - B(k_x^2 + k_y^2)$.\\
Answer: $4 \times4 $  Hamiltonian in momentum space $k_\pm = (k_x + k_y)$ is written in a block diagonal form\\
\[ H(\mathbf{k}) = \begin{pmatrix}
h(\mathbf{k}) & 0\\
0 & h^\ast(\mathbf{k})
\end{pmatrix}\]
where $h(\mathbf{k}) $ is spin up and $h^\ast(\mathbf{-k})$ is spin down related by time reversal  symmetry.\\
The model is defined as
$\vert E, \uparrow \rangle, \vert H, \uparrow \rangle, \vert E, \downarrow \rangle, \vert H, \downarrow \rangle$\\
For spin up block $h(\mathbf{k}):$\\
This block couple: $\vert E, \uparrow \rangle, \vert H, \uparrow \rangle\\$\\

\[ h(\mathbf{k}) = \begin{pmatrix}
 m(\mathbf{k}) & A k_- \\
 A k_+ & -m(\mathbf{k} \\
\end{pmatrix}\]
where: $k\pm = k_x \pm k_y$ then substitute this to $k_- $ and $k_+$. We will get\\
\[ h(\mathbf{k}) = \begin{pmatrix}
 m(\mathbf{k}) & A (k_x - k_y)\\
 A (k_x + k_y) & -m(\mathbf{k}) \\
\end{pmatrix}\]\\
For spin downt this block couple: $\vert E, \downarrow \rangle, \vert H, \downarrow \rangle$\\
\[ h\ast(\mathbf{-k}) = \begin{pmatrix}
 -m(\mathbf{k}) & -A k_+\\
 -A k_- & m(\mathbf{k}) \\
\end{pmatrix}\]\\
Substitute  $k\pm = k_x \pm k_y$ to $k_- $ and $k_+$. We will get\\
\[ h(\mathbf{k}) = \begin{pmatrix}
 -m(\mathbf{k}) & -A (k_x - k_y)\\
 -A (k_x + k_y) & m(\mathbf{k}) \\
\end{pmatrix}\]\\
The given matrix for Hamiltonian is\\
    \[
    H(\mathbf{k}) = \begin{pmatrix}
    m(\mathbf{k}) & A k_x & 0 & -A k_y \\
    A k_x & -m(\mathbf{k}) & -A k_y & 0 \\
    0 & -A k_y & -m(\mathbf{k}) & -A k_x \\
    -A k_y & 0 & -A k_x & m(\mathbf{k})
    \end{pmatrix},
    \]\\
We can write this in spin system matrix\\
    \[
    H(\mathbf{k}) = \begin{pmatrix}
    m(\mathbf{k}) & A k_- & 0 & 0 \\
    A k_+ & -m(\mathbf{k}) & 0 & 0 \\
    0 & 0 & -m(\mathbf{k}) & -A k_+ \\
  0 & 0 & -A k_- & m(\mathbf{k})
    \end{pmatrix},
    \]\\
Why is that the $-A k_y$ in the diagonal becomes 0?\\
$\ast$ The $-A k_y$ does not becomes zero. Since BHZ model does not include spin-flip terms. Primarily due to no coupling between spin-up and spin-down in the BHZ model.

    \item Identify the time-reversal symmetry operator $\Theta$ for this system: $\Theta = \tau_0 \otimes i\sigma_y K$, where $K$ is complex conjugation.\\
Answer: The time reversal operator\\
\[
\Theta = \tau_0\otimes({i\sigma_y})K
\]\\
Where:\\
$\bullet \tau_0: 2\times2$ identity matrix in orbital space\\
$\bullet \sigma_y:$ acts in spin space\\
$\bullet{K}$: Complex conjugation\\
The BHZ Hamiltonian uses the basis\\
\[\vert E, \uparrow \rangle, \vert H, \uparrow \rangle, \vert E, \downarrow \rangle, \vert H, \downarrow \rangle\]\\
Time Reversal Symmetry implies
\[\Theta H({k})\Theta^{-1} = H({-1})
\]\\
We know that the $2\times2$ BHZ Hamiltonian diagonal block is\\
\[ H(\mathbf{k}) = \begin{pmatrix}
h(\mathbf{k}) & 0\\
0 & h^\ast(\mathbf{k})
\end{pmatrix}\]
Where\\
\[ h(\mathbf{k}) = \begin{pmatrix}
 m(\mathbf{k}) & A k_- \\
 A k_+ & -m(\mathbf{k} \\
\end{pmatrix}\]
For spin-up and $h^{\ast}(-k)$ is its time reversed counterpart
Let's apply $\Theta$ and $\Theta^{-1}$ to H(k)\\
\[ \Theta H(k)\Theta^{-1} = (\tau_0\otimes({i\sigma_y}))K (H(k)) K^{-1} (\tau_0\otimes(-{i\sigma_y}) )
\]
\[= (\tau_0\otimes({i\sigma_y})) H^{\ast}(k) (\tau_0\otimes(-{i\sigma_y}))\\ \]
Take the complex conjugate\\
\[ H^{\ast}(\mathbf{k}) = \begin{pmatrix}
h^{\ast}(\mathbf{k}) & 0\\
0 & h(-\mathbf{k})
\end{pmatrix}\]
Then:\\
\[\Theta H^{\ast}(k)\Theta^{-1} = \begin{pmatrix}
\sigma_{y} h^{\ast}(\mathbf{k}) \sigma_{y} & 0\\
0 & \sigma_{y }h(-\mathbf{k}) \sigma_{y}
\end{pmatrix}\]
Using this identity\\
\[\sigma_{y}( d \cdot \sigma)^{\ast} \sigma_{y} = - d \cdot \sigma
\]
Thefore\\
\[\sigma_{y}( h^{\ast}(k)) \sigma_{y} = h(-k)
\]
\[\sigma_{y}( h(-k)) \sigma_{y} = h^{\ast}(-k)
\]So\\
\[ \Theta H(k)\Theta^{-1} = \begin{pmatrix}
h^{\ast}(\mathbf{k}) & 0\\
0 & h(-\mathbf{k})
\end{pmatrix}\]
\[\Theta H(k)\Theta^{-1}= H(-k)\]
Thus, the BHZ Hamiltonian is invariant under time reversal.
    \item Analyze the bulk band structure and demonstrate the topological nature of the insulating phase by calculating the $Z_2$ topological invariant.

    \item Consider a finite-size strip geometry with open boundary conditions.

    \item Analytically derive the dispersion relation of the edge states that appear within the bulk band gap.

    \item Mathematically show how time-reversal symmetry enforces the Kramers degeneracy of the edge states at any momentum $k$ and prevents backscattering between counter-propagating edge states with opposite spin, thus protecting their gaplessness.
\end{enumerate}

\end{document}
